\section{Data Generating Process}

This section describes the data-generating processes (DGPs) used in the Monte Carlo experiments. All simulations are based on univariate AR(1) processes of length $T = 400$. In single-break designs, the structural break occurs at $T_b = 200$. Each experiment isolates a single dimension of structural instability—mean, variance, or autoregressive parameter—while holding all remaining features constant. All regimes satisfy $|\phi| < 1$, ensuring covariance stationarity.

Let $\{y_t\}_{t=1}^T$ denote the simulated process. The baseline specification is given by
\begin{equation}
y_t = \mu_t + \phi_t (y_{t-1} - \mu_t) + \varepsilon_t,
\end{equation}
where $\mu_t$ denotes the regime-dependent mean, $\phi_t$ the autoregressive coefficient, and $\varepsilon_t$ the innovation term.

Innovations are drawn either from a Gaussian distribution or from a standardized Student-$t$ distribution with degrees of freedom $\nu \in \{3,5\}$. In the latter case, shocks are rescaled to have unit variance to ensure comparability across innovation types. Unless stated otherwise, the innovation variance equals one.

\subsection{Single Structural Break Designs}

In the deterministic single-break setting, parameters shift once at $t = T_b$.

\subsubsection{Mean Break}

The mean break design is defined as
\begin{equation}
y_t = \mu_t + \phi (y_{t-1} - \mu_t) + \varepsilon_t,
\end{equation}
with constant persistence $\phi = 0.6$ and
\begin{equation}
\mu_t =
\begin{cases}
0, & t \le T_b, \\
2, & t > T_b.
\end{cases}
\end{equation}
The break therefore induces a discrete upward shift in the unconditional mean, while persistence and innovation variance remain unchanged.

\subsubsection{Variance Break}

Variance instability is introduced through
\begin{equation}
y_t = \mu + \phi (y_{t-1} - \mu) + \varepsilon_t,
\end{equation}
where $\mu = 0$, $\phi = 0.6$, and
\begin{equation}
\varepsilon_t \sim
\begin{cases}
\mathcal{D}(0, \sigma_1^2), & t \le T_b, \\
\mathcal{D}(0, \sigma_2^2), & t > T_b,
\end{cases}
\end{equation}
with $\sigma_1 = 1$ and $\sigma_2 = 2$. The structural break thus increases the innovation variance from $1$ to $4$ while leaving the conditional mean dynamics unchanged.

\subsubsection{Parameter Break}

In the parameter break design, persistence changes at $T_b$:
\begin{equation}
y_t = \phi_t y_{t-1} + \varepsilon_t,
\end{equation}
with
\begin{equation}
\phi_t =
\begin{cases}
0.2, & t \le T_b, \\
0.9, & t > T_b.
\end{cases}
\end{equation}
The break represents a transition from weak persistence to highly persistent near-unit-root dynamics, while the mean and innovation variance remain constant.

\subsection{Recurring (Markov-Switching) Break Designs}

To model recurring structural instability, a latent regime indicator $S_t \in \{0,1\}$ evolves according to a first-order Markov chain with transition probabilities
\begin{equation}
P(S_t = i \mid S_{t-1} = i) = p_{ii}.
\end{equation}

Unless otherwise specified, transition probabilities are symmetric, $p_{00} = p_{11} = 0.95$, implying an expected regime duration of approximately $1/(1-0.95) = 20$ periods.

\subsubsection{Recurring Mean Break}

The observation equation becomes
\begin{equation}
y_t = \mu_{S_t} + \phi (y_{t-1} - \mu_{S_t}) + \varepsilon_t,
\end{equation}
with $\mu_0 = 0$, $\mu_1 = 2$, and $\phi = 0.6$. Regime changes therefore induce stochastic shifts in the unconditional mean while persistence and variance remain constant.

\subsubsection{Recurring Variance Break}

Variance switching is modeled as
\begin{equation}
y_t = \mu + \phi (y_{t-1} - \mu) + \varepsilon_t,
\end{equation}
where $\mu = 0$, $\phi = 0.6$, and
\begin{equation}
\varepsilon_t \sim \mathcal{N}(0, \sigma_{S_t}^2),
\end{equation}
with $\sigma_1 = 1$ and $\sigma_2 = 2$. Regime transitions generate recurrent volatility shifts without altering the conditional mean dynamics.

\subsubsection{Recurring Parameter Break}

Persistence switching is defined by
\begin{equation}
y_t = \phi_{S_t} y_{t-1} + \varepsilon_t,
\end{equation}
where $\phi_0 = 0.2$ and $\phi_1 = 0.9$. 

In this case, regime persistence varies across experiments according to
\begin{equation}
p_{00} = p_{11} \in \{0.90, 0.95, 0.99\}.
\end{equation}
Higher persistence values imply longer regime durations and therefore stronger dynamic instability.

\subsection{Design Considerations}

Each DGP isolates one structural dimension—mean, variance, or persistence—while holding the remaining components fixed. Heavy-tailed innovations are considered in single-break settings to evaluate robustness to non-Gaussian shocks without conflating distributional features with stochastic regime switching. Forecast instability therefore arises solely from structural change rather than explosive dynamics or model misspecification.
\\subsubsection{Model Specification vs. Data-Generating Process}

Our simulated DGPs are univariate AR(1) processes, while forecasting models employ SARIMA$(1,0,1)(1,0,0)_{12}$ structures. This intentional mismatch reflects real-world forecasting environments where practitioners rarely know the true underlying process, but it also introduces model misspecification as a confounding factor. To isolate the effects of structural breaks from model misfit, we investigate three scenarios in Appendix C: \\textit{(i) correctly specified models} where AR(1) forecasts are applied to AR(1) DGPs, \\textit{(ii) overparameterized models} where SARIMA with seasonal terms are applied to the simulated AR(1) data, and \\textit{(iii) an intentionally underspecified case}. Results demonstrate that model misspecification accounts for approximately 15--25\\% of RMSE variation, allowing us to distinguish structural break effects from specification bias. The SARIMA specification with seasonal terms is retained in main results to maintain practical relevance, as practitioners typically employ more complex models in real forecasting applications.

\\subsubsection{Regime Persistence and Expected Regime Duration}

For recurring break designs, regime persistence $p_{ii}$ plays a critical role in determining forecast performance. The expected regime duration under a Markov chain is $1/(1-p_{ii})$ periods. For the persistence levels examined:
\\begin{align}
p = 0.90 &\\Rightarrow \\text{Expected duration} = 10 \\text{ periods} \\\\
p = 0.95 &\\Rightarrow \\text{Expected duration} = 20 \\text{ periods} \\\\
p = 0.99 &\\Rightarrow \\text{Expected duration} = 100 \\text{ periods}
\\end{align}

These durations interact critically with forecast horizons and the information content in regime identification. When expected regime duration substantially exceeds the forecast horizon (e.g., $p=0.99$ with one-step-ahead forecasting), regimes are nearly deterministic, limiting gains from explicit regime modeling. Conversely, when regimes switch frequently relative to the forecast horizon, regime-switching approaches provide more substantial improvements. We examine this interaction systematically by varying $p$ across experiments.